---
title: "PM566_Lab04"
format: 
  html:
    embed resources: true
editor: visual
---

## 1. Read in the data

```{r}
library(tidyverse)
library(data.table)
library(magrittr)
library(leaflet)
library(gganimate)

if (!file.exists("met_all.gz"))
  download.file(
    url = "https://raw.githubusercontent.com/USCbiostats/data-science-data/master/02_met/met_all.gz",
    destfile = "met_all.gz",
    method   = "libcurl",
    timeout  = 60
  )
met <- data.table::fread("met_all.gz")
```

## 2. Prepare the data

Removing temperatures less than -17

```{r}
met <- met[met$temp > -17][elev == 9999.0, elev := NA]
```

Creating date variable

```{r}
met$date_variable <- as.Date(paste(met$year, met$month, met$day, sep = "-"))
```

Keep observations for the first week of the month

```{r}
met$date_week <- data.table::week(as.Date(paste(met$year, met$month, met$day, sep = "-")))
```

Creating an aggregate of mean

```{r}
met_avg <- met[,.(
  temp     = mean(temp,na.rm=TRUE),
  rh       = mean(rh,na.rm=TRUE),
  wind.sp  = mean(wind.sp,na.rm=TRUE),
  vis.dist = mean(vis.dist,na.rm=TRUE),
  dp       = mean(dew.point, na.rm = TRUE),
  lat      = mean(lat),
  lon      = mean(lon), 
  elev     = mean(elev,na.rm=TRUE)
), by=c("USAFID", "day")]
```

Creating a region variable

```{r}
met_avg$regionlat <- cut(met_avg$lat, 
                         breaks=c(24.55525, 39.71, 48.941),
                         labels=c('N', 'S'))
met_avg$regionlon <- cut(met_avg$lon, 
                         breaks=c(-124.29, -98.00, -68.313),
                         labels=c('E', 'W'))
met_avg$region <- paste(met_avg$regionlat, met_avg$regionlon, sep = "")

```

Creating a elevation categorical variable

```{r}
met_avg[, elev_cat := fifelse(elev > 252, "high", "low")]
```

### **3. Use `geom_violin` to examine the wind speed and dew point by region**

```{r}
?geom_violin

met_avg[!is.na(region)] %>% 
  ggplot() + 
  geom_violin(mapping = aes(x = as.factor(1), y = dp, fill = region)) +
  facet_wrap(~ region, nrow = 1)
```

The NE region and NW region has a larger range of dew points than the other regions.

### **4. Use `geom_jitter` with `stat_smooth` to examine the association between dew point and wind speed by region**

```{r}
?stat_smooth
met_avg[!is.na(region)] %>% 
  ggplot(aes(x = dp, y = wind.sp, color = region)) +
  geom_jitter(alpha = 0.5, width = 0.2) +  # Adds jittered points
  stat_smooth(method = "lm", se = FALSE) +  # Adds a linear regression line
  facet_wrap(~ region, ncol = 1)

```

The NE region has a larger distribution of data than the other regions

### **5. Use `geom_bar` to create barplots of the weather stations by elevation category colored by region**

```{r}
ggplot(met_avg, aes(x = met_avg$elev_cat, fill = region)) +
  geom_bar(position = "dodge", na.rm = TRUE) +
  labs(
    title = "Weather Stations by Elevation Category",
    x = "Elevation Category",
    y = "Count"
  ) 

```

There are over 15000 weather stations with low elevation (less than 252) in the NW.

### **6. Use `stat_summary` to examine mean dew point and wind speed by region with standard deviation error bars**

**Dew Point by Region Plot**

```{r}
ggplot(met_avg, aes(x = region, y = dp)) +
  stat_summary(
    fun.data = "mean_sdl", # calculates mean and SD
    fun.args = list(mult = 1),
    geom = "bar",
    width = 0.3,
    fill = "blue",
    color = "black"
  ) +
  stat_summary(
    fun.data = "mean_sdl",
    fun.args = list(mult = 1),
    geom = "errorbar", # creates error bars
    width = 0.1,
    color = "black"
  ) +
  labs(
    title = "Mean Dew Point by Region",
    x = "Region",
    y = "Dew Point"
  ) 
```

NW has the largest mean dew point, and NE has the most variability (largest SD) for dew point.

**Wind Speed by Region**

```{r}
ggplot(met_avg, aes(x = region, y = wind.sp)) +
  stat_summary(
    fun.data = "mean_sdl", # calculates mean and SD
    fun.args = list(mult = 1),
    geom = "bar",
    width = 0.3,
    fill = "blue",
    color = "black"
  ) +
  stat_summary(
    fun.data = "mean_sdl",
    fun.args = list(mult = 1),
    geom = "errorbar", # creates error bars
    width = 0.1,
    color = "black"
  ) +
  labs(
    title = "Mean Wind Speed by Region",
    x = "Region",
    y = "Wind Speed"
  ) 
```

NE has the largest mean for wind speed, and SE has the largest variability in wind speed data (highest SD).

### **7. Make a map showing the spatial trend in relative humidity in the US**

```{r}
temp.pal <- colorNumeric(c('lightgreen','pink','purple'), domain = met_avg$rh)

rh_map <- leaflet(met_avg) %>%
  addProviderTiles('CartoDB.Positron') %>%
  addCircles(
    lat = ~lat,
    lng = ~lon,
    label = ~paste0(round(rh, 2), ' % RH'),  # Label with RH values
    color = ~temp.pal(rh),  # Use the RH palette
    opacity = 1,
    fillOpacity = 1,
    radius = 500
  ) %>%
  addLegend(
    'bottomleft',
    pal = temp.pal,
    values = met_avg$rh,
    title = 'Relative Humidity (%)',
    opacity = 1)

  
# top 10 rank but it doesn't work 

# top_10 <- met_avg$rh %>%
#   arrange(desc(rh)) %>%
#   head(10)

    # addMarkers(
    # data = top_10,
    # popup = ~paste("RH:", rh, "%"),
    # label = ~paste("Rank:", rank(-rh)),
    # color = ~custom_colors(rh),
    # labelOptions = labelOptions(noHide = TRUE)
    
rh_map
```

Relative humidity seems to be higher on the east coast than the west coast

### **8. Use a ggplot extension**

```{r}
simple_plot <- ggplot(met_avg, aes(x = met_avg$elev_cat, fill = region)) +
  geom_bar(position = "dodge", na.rm = TRUE) +
  labs(
    title = "Weather Stations by Elevation Category",
    x = "Elevation Category",
    y = "Count"
  ) +
  transition_states(
    gear,
    transition_length = 2,
    state_length = 1
  ) +
  enter_fade() + 
  exit_shrink() +
  ease_aes('sine-in-out')
simple_plot

```
